Fasano--Franceschini Test: an Implementation of a 2-Dimensional Kolmogorov--Smirnov in \proglang{R}

### 2021-06-08 journalMS/article.tex
updated minor edits

%% - \Abstract{} almost as usual
\Abstract{
\RBnote{Perhaps briefly state what the KS test is.}
The versatility of the univariate Kolmogorov-Smirnov (KS) test has
made it a cornerstone of statistical analysis and is commonly used
across the scientific disciplines. However, the test proposed by
Kolmogorov and Smirnov does not naturally extend to multidimensional
distributions.  Here, we present the \pkg{fasano.franceschini.test}
package, an  \proglang{R} implementation of the 2-D KS two--sample
test as defined by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides three improvements
over the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; (ii) the package implements a procedure for handling ties in
the data; and (iii) the package implements a parallelized
bootstrapping procedure for improved significance testing. Ultimately,
the \pkg{fasano.franceschini.test} package presents a robust
statistical test for analysing random samples defined in 2-dimensions.
}

@@

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered in each point of the given samples to compute the
cumulative distribution functions.  Rather than looking over all 
$X_i, Y_j: i,j={1,n}$, Fasano and Franceschini only used the detected points
$X_i, Y_i : i={1,n}$. 
\RBnote{I don't follow the preceding line? (Notation?)}
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

%\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the points from \textit{other} sample
as the origins to compute the maximal $D_{FF,2}$ with origins from the
second sample. $D_{FF,1}$ and $D_{FF,2}$  are then averaged to compute
the overall $D_{FF}$ for hypothesis testing. \RBnote{clarify: is
$D_{FF}=(D_{FF,1}+D_{FF,2})/2$, or is it weighted by the $n$?}

@@

Using Monte Carlo simulation, Fasano and Franceschini created a
look-up table of critical values of $D_{FF}$ as a function of
$D_{FF}$, the sample size, and the coefficient of correlation $r$.
\cite{numericalRecipes} later defined an approximate fit to the lookup table
as follows. 
%
For the a single sample of size $n$,
\begin{equation} \label{eq:4}
\mathbb{P}(d_{FF} > D_{FF}) = \Phi \left( \frac{D_{FF}\sqrt{n}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{n})} \right) \, .
\end{equation}
\RBnote{Is $\Phi$ the same function defined in Eq~\ref{eq:1}?}
The two sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
n = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined in the usual way as
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})}{\sqrt{\sum_{i}(X_i-\bar{X})^2}\sqrt{\sum_{i}(Y_i-\bar{Y})^2}}\, .
\end{equation}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
The univariate Kolmogorov-Smirnov (KS) test is a non--parametric
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample
case,  whether the two
samples come from the same underlying distribution).
The versatility of the KS test has
made it a cornerstone of statistical analysis and is commonly used
across the scientific disciplines. However, the test proposed by
Kolmogorov and Smirnov does not naturally extend to multidimensional
distributions.  Here, we present the \pkg{fasano.franceschini.test}
package, an  \proglang{R} implementation of the 2-D KS two--sample
test as defined by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides three improvements
over the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; (ii) the package implements a procedure for handling ties in
the data; and (iii) the package implements a parallelized
bootstrapping procedure for improved significance testing. Ultimately,
the \pkg{fasano.franceschini.test} package presents a robust
statistical test for analysing random samples defined in 2-dimensions.
}

@@

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered in each point of the given samples to compute the
cumulative distribution functions.  Rather than looking over all
$X_i, Y_j \,;\; i,j \in [1,n]$, Fasano and Franceschini only used the detected points
$X_i, Y_i \,;\; i \in [1,n]$.
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

%\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the points from \textit{other} sample
as the origins to compute the maximal $D_{FF,2}$ with origins from the
second sample. $D_{FF,1}$ and $D_{FF,2}$  are then averaged to compute
the overall $D_{FF}$ for hypothesis testing -- e.g.
$D_{FF}=(D_{FF,1}+D_{FF,2})/2$.

@@

Using Monte Carlo simulation, Fasano and Franceschini created a
look-up table of critical values of $D_{FF}$ as a function of
$D_{FF}$, the sample size, and the coefficient of correlation $r$.
\cite{numericalRecipes} later defined an approximate fit to the lookup table
as follows.
%
For the a single sample of size $n$,
\begin{equation} \label{eq:4}
\mathbb{P}(d_{FF} > D_{FF}) = \Phi \left( \frac{D_{FF}\sqrt{n}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{n})} \right) \, .
\end{equation}
where $\Phi$ is the same function defined in Eq~\ref{eq:1}.
The two sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
n = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined in the usual way as
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})}{\sqrt{\sum_{i}(X_i-\bar{X})^2}\sqrt{\sum_{i}(Y_i-\bar{Y})^2}}\, .
\end{equation}
